---
title: "Walkthrough of shinymrp workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Walkthrough of shinymrp workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette provides more details about the individual methods of the [`MRPWorkflow`][MRPWorkflow] object, including their purposes, input requirements, and caveats. For a general overview, please refer to the ["Getting started with shinymrp"](https://mrp-interface.github.io/shinymrp/articles/getting-started.html) vignette.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}

ggplot2::theme_update(
  text = ggplot2::element_text(size = 12),
  plot.margin = ggplot2::margin(0, 0, 0, 0)
)
```

## Initializing the workflow

The process begins with creating an `MRPWorkflow` object using the `mrp_workflow` function. The object-oriented design allows the object to manage the essential data under the hood and offer a clean interface that does not require the users to input the same piece of information repeatedly.

```{r}
library(shinymrp)

workflow <- mrp_workflow()
```

## Data preparation

The goal of this stage is to produce two essential pieces of data:

- **Sample data**: The analysis sample that includes the outcome of interest (either binary or continuous ) and predictors, such as the COVID test records and survey sample results.
- **Poststratification data**: The table containing sizes of groups in the target population defined by the demographic and geographic factors.

### Use cases

```{r}
sample_data <- example_sample_data(
  is_timevar = TRUE,
  is_aggregated = TRUE,
  special_case = NULL,
  family = "binomial"
)
```

### Individual-level vs. Aggregated Data

The interface accepts data in two formats:

- **Individual-level**: Each row contains information for on individual.
- **Aggregated**: Each row contains information for one group (e.g., White males aged 18-30 in Michigan), with geographic-demographic factors, total numbers of individuals, and summary of outcomes.

Data with continuous outcome measures are expected only at individual-level. For data with binary outcome measures, the aggregated format is preferred for computational benefits. Individual-level data will be automatically aggregated upon upload. Data requirements vary slightly between formats, mainly regarding the outcome measure.

Required Columns and Categories
The application screens input data using a specific naming convention. Here's a list of the expected columns and their values (case-insensitive):

- Sex: male, female
- Race: Black, White, other
- Age
- Education attainment (edu): below high school (no hs), high school (hs), some college, 4-year college, post-grad
- ZIP code [^geo]
- County [^geo]
- State [^geo]
- Time indices (time) [^time]
- Date
- Continuous outcome measure (outcome) [^outcome-cont]
- Positive response indicator or number of positive responses (positive) [^outcome-bin]
- Cross-tabulation cell counts (total) [^outcome-bin]
- Survey weights (weight) [^weight]

[^geo]: For general use cases, providing geographic information is optional. The application will automatically identify the smallest geographic scale available and provide the corresponding higher levels.
[^time]: If the input sample data are in aggregated format, there has to be a column named 'time' that contains time indices. An optional 'date' column containing the date of the first day of each period can be included for visualization purposes. For individual-level sample data, the interface will automatically convert the dates to time indices, but users can also provide the time indices directly. The interface uses time-invariant poststratification data.
[^outcome-cont]: For data with continuous outcome measures, the outcome column must be named 'outcome'.
[^outcome-bin]: For binary outcome measures, the outcome column of individual-level data must be named 'positive'. Aggregated data require two columns to represent the outcome measures: the total count of individuals and the number of positive responses for each cross-tabulation cell, which should be named 'total' and 'positive', respectively.
[^weight]: Please name the column containing survey weights in the data 'weight'. If the uploaded poststratification data include survey weights, the interface uses weights to estimate the population counts.

```{r}
workflow$preprocess(
  sample_data,
  is_timevar = TRUE,
  is_aggregated = TRUE,
  special_case = NULL,
  family = "binomial"
)
```

### Linking data to the ACS
```{r}
workflow$link_acs(link_geo = "zip", acs_year = 2021)
```

### Load custom poststratification data
```{r}
pstrat_data <- example_pstrat_data()
workflow$load_pstrat(pstrat_data, is_aggregated = TRUE)
```

## Visualizing descriptive statistics

```{r}
workflow$demo_bars(demo = "sex")
```

```{r}
workflow$sample_size_map()
```

```{r}
workflow$outcome_plot()
```

```{r}
workflow$outcome_map()
```

## Model fitting and diagnostics


### Model specification

```{r}
model1 <- workflow$create_model(
  model_spec = list(
    Intercept = list(
      Intercept = ""
    ),
    fixed = list(
      sex = "",
      race = ""
    ),
    varying = list(
      age = "",
      time = ""
    )
  )
)
```

```{r}
model2 <- workflow$create_model(
  model_spec = list(
    Intercept = list(
      Intercept = ""
    ),
    fixed = list(
      sex = "",
      race = ""
    ),
    varying = list(
      age = "",
      time = ""
    ),
    interaction = list(
      `age:time` = "",
      `race:time` = "",
      `sex:time` = ""
    )
  )
)
```

### Run MCMC

```{r}
model1$fit(n_iter = 500, n_chains = 2, seed = 123)
model2$fit(n_iter = 500, n_chains = 2, seed = 123)
```

### Diagnostics

```{r}
model1$diagnostics()
```

```{r}
model1$summary()
```

```{r}
workflow$pp_check(model1)
```

```{r}
model2$diagnostics()
```

```{r}
model2$summary()
```

```{r}
workflow$pp_check(model2)
```

```{r, compare_models}
workflow$compare_models(model1, model2)
```

### Saving and loading models

```{r, save_model, eval=FALSE}
workflow$save_model(model1, file = "model1.qs")
```

```{r, load_model, eval=FALSE}
model1 <- workflow$load_model("model1.qs")
```


## Visualize estimates

```{r, estimate_plot, fig.height=10}
workflow$estimate_plot(model1, group = "sex")
```

```{r, estimate_map}
workflow$estimate_map(model1, geo = "county")
```
